Finally Tagless, Partially Evaluated

### 2007-06-08 tagless/tagless-final.tex
Retarget for APLAS and JFP in parallel (the text is completely the same)

\listfiles
\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
\usepackage{mdwlist}
\usepackage{mdwtab}
\usepackage{dblfloatfix}

% Set pdftex paper size
\ifx\pdfoutput\undefined\else \ifx\pdfoutput\relax\else \ifcase\pdfoutput\else
    \setlength{\pdfpagewidth}{8.5in}
    \setlength{\pdfpageheight}{11in}
\fi \fi \fi

\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
\newcommand{\ccshan}[1]{{\it [Ken says: #1]}}

\usepackage{natbib1}
\let\cite=\citep

\usepackage[compact]{fancyvrb1}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\parindent}

\newtheorem{prop}{Proposition}

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\BB}{\mathbb{B}}

\newcommand{\fun}[1]{\mathopen{\lambda\mathord{#1}.\,}}
\newcommand{\fix}[1]{\mathopen{\mathrm{fix\,}\mathord{#1}.\,}}
\newcommand{\Forall}[1]{\mathopen{\forall\mathord{#1}.\,}}
\newcommand{\Exists}[1]{\mathopen{\exists\mathord{#1}.\,}}
\newcommand{\cond}[3]{\mathrm{if\ }#1\mathrm{\ then\ }#2\mathrm{\ else\ }#3}
\newcommand{\be}[1]{\mathrm{let\ }#1\mathrm{\ in\ }}
\newcommand{\True}{\mathinner{\mathrm{true}}}
\newcommand{\False}{\mathinner{\mathrm{false}}}
\newcommand{\Encode}[1]{\mathinner{\mathopen{\text{``}}#1\mathclose{\text{''}}}}
\newcommand{\encode}[1]{\mathinner{\mathopen{\text{`}}#1\mathclose{\text{'}}}}
\newcommand{\ident}[1]{\mathinner{\text{\textit{#1\/}}}}
\newcommand{\subst}[3]{#1\left\{#2\mapsto#3\right\}}
\DeclareMathOperator{\pe}{PE}
\DeclareMathOperator{\si}{SI}

% Make space around section headings no less than space around theorems
\makeatletter
\@tempdima\topsep \advance\@tempdima\parsep
\ifdim\@sectionbelowskip   <\topsep \@sectionbelowskip   \topsep \fi
\ifdim\@subsectionbelowskip<\topsep \@subsectionbelowskip\topsep \fi
\makeatother

% Allow things to flow a little more tightly around figures
\renewcommand\floatpagefraction{.95}
\renewcommand\topfraction{.95}
\renewcommand\bottomfraction{.95}
\renewcommand\textfraction{.05}

% Match the rule inserted by \@makecaption in sigplanconf.cls
\setproofrulebreadth .33pt
\newenvironment{floatrule}
    {\hrule width \hsize height .33pt \vspace{.5pc}}
    {\par\addvspace{1ex}}

% Make some Rel symbols into Bin symbols instead
\DeclareMathSymbol{\to}{\mathbin}{symbols}{"21}
\DeclareMathSymbol{:}{\mathbin}{operators}{"3A}

\begin{document}

% http://flint.cs.yale.edu/aplas2007/

\conferenceinfo{ICFP '07}{Freiburg, Germany} 
\copyrightyear{2007} 
\copyrightdata{[to be supplied]}

\makeatletter
\titlebanner{Submitted to \@conferencename}        % These are ignored unless
\preprintfooter{\rlap{\@titletext}}   % 'preprint' option specified.
\def \@formatyear {\llap{\number\year/\number\month/\number\day}}
\makeatother

\title{Finally Tagless, Partially Evaluated}
\subtitle{Tagless Staged Interpreters for Simpler Typed Languages}
%Staged typed type-preserving tagless ``interpreters'', \emph{finally}

\begin{comment}
\authorinfo{Jacques Carette}
           {McMaster University}
           {carette@mcmaster.ca}
\authorinfo{Oleg Kiselyov}
           {FNMOC}
           {oleg@pobox.com}
\authorinfo{Chung-chieh Shan}
           {Rutgers University}
           {ccshan@cs.rutgers.edu}
\end{comment}
\authorinfo
{\hspace*{-15em}\smash{\normalsize\it\begin{minipage}[t]{30em}
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm\citep{reynolds-definitional}
\end{minipage}}\hspace*{-15em}}{}{}

\maketitle

@@

@@

Comparing |Symantics| with Figure~\ref{fig:object}
shows how to represent every typed, closed object term in the
metalanguage. Moreover, the representation preserves types, in the following
sense.
\begin{prop}
If an object term has the object type~$t$, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr |$t$.
\end{prop}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

@@

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{prop}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{prop}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{prop}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{prop}
These propositions follow immediately from the soundness of the
metalanguage's type system.

@@

Instead of adding higher-rank and higher-kind polymorphism to our object
language (along with polymorphism over kinds!), we add let-bound polymorphism.
As usual,
we can add a new typing rule
\begin{equation*}
    \begin{prooftree}
        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
        \justifies \be{x=e_1} e_2 : t_2
    \end{prooftree}
    .
\end{equation*}
The pre-encoding of a let\hyp expression is trivial.
\begin{align*}
    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
hole may occur under a binder, so plugging a term into the context may
capture free variables of the term.  By pre-encoding a hole to a hole,
we extend pre-encoding from a translation on terms to one on
contexts.}
    \encode{[~]} \quad &= \quad [~]
\end{align*}
We define an interpreter in the object language to be not a term but
a context.  For example, the evaluator is the context
\begin{align*}
    \si[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} x&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
        \be{\ident{lam} &= \fun{f} f&} \\
        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
    \end{tabular}
\intertext{and the size\hyp measurer is the context}
    \mathrm{SZ}[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} 1&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
    \end{tabular}
\end{align*}
To interpret an object term~$e$ using an interpreter $I[~]$ is to
evaluate the object term~$I[\encode{e}]$.  $\si$ is
a self\hyp interpreter in the following sense.
\begin{prop}
    $\si[\encode{e}]$ is observationally equivalent to~$e$.
\end{prop}
As a corollary, we can pre-encode the self\hyp interpreter itself as
a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
$\si$ can interpret itself.  Our partial evaluator is optimal with respect to
the self\hyp interpreter~$\si$.
\begin{prop}
    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
    either both undefined or both defined and
    equal up to $\alpha$\hyp conversion.
\end{prop}

@@

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
kindly provided some useful references on coalgebraic structures in the
$\lambda$\hyp calculus. We thank anonymous reviewers for pointing out
related work.

>>>>

\begin{comment}
\authorinfo{Jacques Carette}
           {McMaster University}
           {carette@mcmaster.ca}
\authorinfo{Oleg Kiselyov}
           {FNMOC}
           {oleg@pobox.com}
\authorinfo{Chung-chieh Shan}
           {Rutgers University}
           {ccshan@cs.rutgers.edu}
\end{comment}

@@

\begin{quote}
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm\citep{reynolds-definitional}
\end{quote}

@@

Comparing |Symantics| with Figure~\ref{fig:object}
shows how to represent every typed, closed object term in the
metalanguage. Moreover, the representation preserves types, in the following
sense.
\begin{proposition}
If an object term has the object type~$t$, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr |$t$.
\end{proposition}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

@@

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
These propositions follow immediately from the soundness of the
metalanguage's type system.

@@

Instead of adding higher-rank and higher-kind polymorphism to our object
language (along with polymorphism over kinds!), we add let-bound polymorphism.
As usual,
we can add a new typing rule
\begin{equation*}
    \begin{prooftree}
        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
        \justifies \be{x=e_1} e_2 : t_2
    \end{prooftree}
    .
\end{equation*}
The pre-encoding of a let\hyp expression is trivial.
\begin{align*}
    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
hole may occur under a binder, so plugging a term into the context may
capture free variables of the term.  By pre-encoding a hole to a hole,
we extend pre-encoding from a translation on terms to one on
contexts.}
    \encode{[~]} \quad &= \quad [~]
\end{align*}
We define an interpreter in the object language to be not a term but
a context.  For example, the evaluator is the context
\begin{align*}
    \si[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} x&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
        \be{\ident{lam} &= \fun{f} f&} \\
        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
    \end{tabular}
\intertext{and the size\hyp measurer is the context}
    \mathrm{SZ}[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} 1&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
    \end{tabular}
\end{align*}
To interpret an object term~$e$ using an interpreter $I[~]$ is to
evaluate the object term~$I[\encode{e}]$.  $\si$ is
a self\hyp interpreter in the following sense.
\begin{proposition}
    $\si[\encode{e}]$ is observationally equivalent to~$e$.
\end{proposition}
As a corollary, we can pre-encode the self\hyp interpreter itself as
a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
$\si$ can interpret itself.  Our partial evaluator is optimal with respect to
the self\hyp interpreter~$\si$.
\begin{proposition}
    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
    either both undefined or both defined and
    equal up to $\alpha$\hyp conversion.
\end{proposition}

@@

\section*{Acknowledgments}
\ourthanks
